---
title: "Customer segmentation for products marketing"
author: "Reagan Kesseku"
date: "2023-01-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
# set global chunk options
# -------------------------
knitr::opts_chunk$set(echo = TRUE, cache = F, tidy = T,
                      warning = F, message = F, options(digits = 3))
```

```{r}
# set working directory
# ---------------------
setwd("D:/Ph.D_materials/Programming/R_programming/mdsr/customer-segmentation_analysis")
```

```{r}
# Load functions and packages
# ---------------------------
source("pkg.R")
```

## DATA VISUALIZATION 

```{r}
# Import the Mall customers data 
# --------------------------------
customers <-  vroom::vroom("Mall_Customers.csv", col_names =  T)

customers_old <- customers

# Take a glimpse of the data sets
customers %>% glimpse()

# Change gender to factor 
# -----------------------
customers <- customers %>% mutate(
  Gender = factor(Gender)
)



# rename income and spending variables
# ------------------------------------
customers <- customers %>% rename(
  annual_income = 'Annual Income (k$)',
  spending_score = 'Spending Score (1-100)'
  )

```

There are `r dim(customers)[1]` observations and `r dim(customers)[2]` variables in the movies data. Additionally, all variables were numerical. However, we convert the class variables to factors.


#### EXPLORATORY DATA ANAYSIS

```{r}
# check the five number summary and other measures of Amount
# ----------------------------------------------------------
d <- favstats(Age~ Gender , data = customers)

knitr::kable(d, digits = 3, 
             format.args = list(scientific = FALSE),
              caption = 'Descriptive summary of age by gender.')

```


```{r , gender-customer-dist}
tt <- ggplot(data = customers, aes(x = Gender)) + 
  geom_bar(aes(fill = Gender)) +
  labs(y = "Number of values in class", 
       title = "Bar graph of the target variable class") +
  theme_bw() 
```
Clearly, this shows a highly imbalanced classification problem.

```{r "density-plot-Annual_income-gender"}
# calculate median income and spending across gender 
mu_income <- plyr::ddply(customers, "Gender", summarise, grp.median = median(annual_income))

mu_spend <- plyr::ddply(customers, "Gender", summarise, grp.median = median(spending_score))

# plot graph
# -----------
 p1 <- ggplot(data = customers, aes(x = annual_income, color = Gender, fill = Gender)) +
  geom_density(alpha=0.3) +
  geom_vline(data = mu_income, aes(xintercept = grp.median, color = Gender),
             linetype = "dashed") +
 scale_fill_manual(values = c("#E7B800", "#00AFBB"))+
  labs(title = "Annual income distribution across gender",
       x = "Annual income") +
  theme_bw()


 p2 <- ggplot(data = customers, aes(x = spending_score, color = Gender, fill = Gender)) +
  geom_density(alpha=0.3) +
  geom_vline(data = mu_spend, aes(xintercept = grp.median, color = Gender),
             linetype = "dashed") +
 scale_fill_manual(values = c("#E7B800", "#00AFBB"))+
  labs(title = "Spending score across gender",
       x = "Spending score") +
  theme_bw()

p1 / p2
```

```{r, density }
# Change violin plot colors by gender
# ------------------------------------
p <- ggplot(data = customers, aes(x = Gender, y = annual_income)) +
  geom_violin(trim = FALSE, aes(fill = Gender)) +
  geom_boxplot(width = 0.15, color = c("blue", "green"),
               fill = c("white", "white")) + 
  geom_point() + 
  labs(y = "Annual income of customers", 
       title = "Violin plot of income distribution") +
  theme_bw()
p
```

```{r}
customers %>% select(Gender) %>% unique()


# scale spending and income variable
# -----------------------------------
customers <- customers %>% mutate(
  annual_income = scale(annual_income),
  spending_score = scale(spending_score)
  )
```
The descriptive statistics show that the amount values are highly variable. 
This suggests the we scale the data as it helps with most machine learning algorithms.

```{r}
#Elbow Method for finding the optimal number of clusters
set.seed(123)
# Compute and plot wss for k = 2 to k = 15.
k.max <- 10
kk =  customers %>% select(annual_income, spending_score)
data <- kk
wcss <- sapply(1:k.max,
        function(k){kmeans(data, k, nstart= 10,iter.max = 350)$tot.withinss}
        )
wcss
kt <- data.frame("k.max" = 1:k.max, "wcss" = wcss)

# plot graph
ggplot(kt, aes(x = k.max, y = wcss)) +
  geom_line(color = "red") +
  geom_point() +
labs(y = "Within-cluster sum of squares",
     x = "Number of clusters",
     title = "Using \"Elbow method\" to choose appropriate K") +
  theme_bw()

```

```{r}
set.seed(6)
library(cluster)

clust_Variables <- customers %>% select(annual_income, spending_score) %>% kmeans(5, iter.max = 300, nstart = 2) %>% fitted("classes") %>% as.factor()


customers <- customers %>%
mutate(clust_Variables = clust_Variables)


p3 <- customers %>% ggplot(aes(x = annual_income, y = spending_score)) +
  geom_point(aes(color = clust_Variables), alpha = 0.5) + 
  scale_color_brewer(palette = "Set2") + 
  labs(y = "Spending score of customers",
       x = "Annual income of customers",
       title = "K-Means clustering of customers income and spending") +
  theme_bw()


p4 <- customers %>% ggplot(aes(x = annual_income, y = spending_score)) +
  geom_point(aes(color = clust_Variables), alpha = 0.5) + 
  scale_color_brewer(palette = "Set2") + 
  labs(y = "Spending score of customers",
       x = "Annual income of customers",
       title = "K-Means clustering of customers income and spending by Gender") +
  facet_wrap( ~ Gender, nrow = 1) +
  theme_bw()
p3 / p4
```


```{r}
# selecting cluster 4 
 customers %>% select(CustomerID) %>%
  filter(clust_Variables == "4") %>% as.vector()
```


